\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2022-09-29 sections/results.tex
Change: Add and update references for the introduction
Update on Overleaf.

  We developed \ac{micone}, a flexible and modular pipeline for the inference of co-occurrence networks from 16S data.
  \ac{micone} incorporates various popular, publicly available tools as well as custom Python modules for 16S data analysis and network inference (see Methods).
  The different steps that are a part of the \ac{micone} co-occurrence network inference workflow (Figure~\ref{fig:figure1}) can be grouped into five major modules; (i) \ac{sp}; (ii) \ac{dc}; (iii) \ac{ta}; (iv) \ac{op}; (v) \ac{ni}.
  Each process in the pipeline can be implemented through  multiple tools (see Methods and Figure~\ref{fig:figure1}).
  The effects of changing any intermediate step of the pipeline can be evaluated in terms of the final network outcome, as well as on any of the intermediate metrics and data outputs.
  The choice of tools and parameters is encoded in a configuration file (with defaults shown in Tables S2-S6 at \href{https://github.com/segrelab/MiCoNE-pipeline-paper}{https://github.com/segrelab/MiCoNE-pipeline-paper}).
  Through a systematic analysis of tool combinations at each step of the pipeline, we estimated how much the final co-occurrence network depends on the possible choices at each step.

  Our analysis involves two types of data: The first type consists of sets of 16S sequencing data from samples of human stool microbiomes from a fecal microbiome transplant (FMT) study of autism~\cite{Kang2017}.
  The second type is a collection of datasets synthetically or artificially created for the specific goal of evaluating computational analysis tools.
  In particular, in order to  compare how well each step in \ac{micone} best captures the underlying data, we use both mock data (labeled mock4, mock12, and mock16) from mockrobiota~\cite{Bokulich2016} and synthetic networks generated using the NorTA~\cite{Kurtz2015} and seqtime~\cite{Rottjers2018} approaches (See Methods).

@@

  \subsection*{DC: Denoising and clustering differ in their identification of sequences that are low in abundance}

  The \ac{dc} step is commonly carried out to generate representative sequences (in the form of the \acs{otu}/\acs{esv} tables) from the demultiplexed and trimmed 16S sequencing data.
  In order to compare the count tables generated by different tools, we processed the same 16S sequencing reads (from the FMT study~\cite{Kang2017}) using 5 different methods: open-reference clustering, closed-reference clustering, de novo clustering, \ac{dada2}~\cite{Callahan2016} and Deblur~\cite{Amir2017}.
  The first three methods are from the vsearch plugin from \ac{qiime2}~\cite{bolyenReproducibleInteractiveScalable2019}.
  The closed and open reference methods in this analysis use the \ac{gg}~\cite{DeSantis2006} database for reference sequence alignment.

  A comparison of the different approaches is visualized in the form of mean UniFrac values across all samples (Figure~\ref{fig:figure2}).
  The analysis was performed using both the weighted UniFrac~\cite{Lozupone2007} (Figure~\ref{fig:figure2}A) distance metric, which takes into account the counts of the representative sequences, and the unweighted UniFrac~\cite{Lozupone2005} (Figure~\ref{fig:figure2}B) distance metric, which gives equal weights to each sequence.

  The first main message emerging from this analysis is that the representative sequences generated by the different methods, with the exception of Deblur, are similar to each other when weighted by their abundance (Figure~\ref{fig:figure2}A).
  A second message is that the different methods differ mainly in the assignment of sequences of lower abundance. This can be inferred from the fact that the unweighted comparison (Figure~\ref{fig:figure2}B) shows an increase in dissimilarity between each pair of methods (see additional details in Supplementary and Figure \ref{fig:figure_s3}).

  These comparisons only elucidate the similarity of a pair of methods.
  In order to determine which tool most accurately recapitulates the reference sequences in the samples, we applied the same pipeline step to process the mock datasets (mock4, mock12, and mock16) and compared the predicted representative sequences with the true sequences and their distribution.
  The results (Figure~\ref{fig:figure2}C and D) show that the predicted sequence distributions are overall considerably different from the expected ones.
  The significant variation across datasets indicates that the datasets themselves play a big role in method performance.
  There is no method that outperforms the rest in all datasets (see Supplementary for an extended discussion).
  Based on their slightly better performance on the mock datasets, their de novo error correcting nature and previous independent evaluation~\cite{Nearing2018}, \ac{dada2} and Deblur appear to be the most reliable.
  This is because the open-reference and de novo clustering methods return a much larger number of \ac{otu}s compared to the other pipelines and this would affect the accuracy of the network inference step if filtering is not performed.
  Overall, since \ac{dada2} as compared to Deblur, displays better performance on all the mock datasets on the weighted UniFrac metric, we set this tool as the default for the DC step of the pipeline.
  However, if comparison across studies that have sequenced different 16S regions is required, closed-reference and open-reference might be a better option.

  After the denoising, the sequences are subject to chimera checking.
  The \ac{micone} pipeline supports two different chimera checking methods, ``uchime-denovo"~\cite{bolyenReproducibleInteractiveScalable2019}, and ``remove bimera"~\cite{Callahan2016}.
  We did not notice any significant difference between the two methods (Figure~\ref{fig:figure_s4}), implying that they identify and remove mostly the same set of sequences as chimeras.
  Since the remove bimera method was originally developed in conjunction with dada2 we use this method as the default for the chimera checking module.
  The DC step thus results in a reduced set of unique sequences, which will be referred to as representative sequences in the subsequent steps.

@@

  After assigning all the representative sequences to taxonomies we performed a pairwise comparison of the similarity between the top 100 assignments (by abundance) from different databases at every taxonomic level (Figure~\ref{fig:figure3}B).
  The comparisons of the assignments beyond the Family level (Family, Genus, and Species) show less than $45\%$ similarity between any pair of databases.
  This implies that the taxonomy assignments from each reference database are fairly unique.
  The comparison of all assigned genera (Figure~\ref{fig:figure_s5}), instead of just the top 100, contains a higher percentage of mismatches.
  This suggests that many of the most abundant sequences are consistently matched to the same taxonomies in the different reference databases, at least for the dataset tested in the current analysis.

@@

  Since no database performs better than others against mock datasets, the choice of which database to use could be driven by other reasons (see Supplementary discussion).
  One reason to choose a particular database could be the frequency of updates and the potential for future growth.
  Both \ac{gg}, due to its frequent use in the literature~\cite{Balvociute2017}, and \ac{ncbi}, due to its regular revision and maintenance, could be good choices for taxonomy assignment, with few differences observed between the two in our analysis  (Figure~\ref{fig:figure_s5}).
  In our default pipeline, we choose \ac{gg} as the default method.

@@

  Similar to what was done for the previous steps of the pipeline, and in analogy with previous estimations of network inference accuracy~\cite{Kurtz2015,Weiss2016}, we evaluated the network inference algorithms and the final consensus network using synthetic interaction data.
  For this purpose, we generated synthetic interaction data using the ``NorTA''~\cite{Kurtz2015} and ``seqtime''~\cite{faustSignaturesEcologicalProcesses2018} methods (see Methods).
  For each method, an \ac{otu} counts table is generated based on the selected interaction network topology and properties.
  This counts table is used as the input to the \ac{micone} pipeline to generate predicted associations.
  The interaction network used to generate the counts table is used as the source of true interactions to calculate the precision (Figure \ref{fig:figure5}) and sensitivity (Figure \ref{fig:figure_s6} and Figure \ref{fig:figure_s7}) for each network inference algorithm.
  As shown in Figure \ref{fig:figure5} the consensus algorithm, especially the scaled-sum method, captures true associations with high precision (through the removal of edges that are either not present in most of the inference methods or whose association strength is low across methods).
  Overall, the scaled-sum method for $p=1.000$ performs the best (precision = $1.000$ for both NorTA and seqtime).
  The scaled-sum method for $p=0.333$ (default option in the pipeline) shows a high precision ($0.956$ with NorTA; $0.688$ with seqtime), without displaying significant reduction in sensitivity (Figure~\ref{fig:figure_s6} and Figure~\ref{fig:figure_s7}).
  However, if higher precision is required $p>0.5$ can be considered.
  Therefore, the consensus networks provide the means to obtain a short list of associations that would have a high probability of being present in the real association network.

@@

  In order to analyze the effect of different processing methods on the inferred co-occurrence networks (before consensus estimation), we generated networks using all possible combinations of methods and quantified the variability due to each choice (Figure \ref{fig:figure6}A).
 This was achieved by building a linear model of the edges of the network as a function of the various steps in the pipeline workflow (see Methods).
  Figure \ref{fig:figure6}A, shows the percentage of total variation among the co-occurrence networks due to the different steps of the pipeline.
  The \ac{ta} step, or more specifically the choice of 16S reference database, contributes the most ($\sim65.4\%$) to the variation in the networks, followed by the \ac{op} step ($\sim26.8\%$).
  This result highlights the importance of the taxonomy assignment step in the 16S data analysis workflow, implying that a change in the reference database will result in drastically different inferred networks.
  This is likely due to the differential assignment of representative sequences to taxonomic entities (Figure~\ref{fig:figure3} and Figure~\ref{fig:figure_s5}), which drastically alter the nodes and hence the underlying network topology.

  The effects of the different steps of the pipeline on the inferred networks can be best visualized through dimensionality reduction.
  The PCA in Figure \ref{fig:figure6}B shows all the above networks, colored by the databases used in TA (left) and by the NI approach (right).
  The major effect of the TA step choice, shown before in Figure \ref{fig:figure6}A, is also reflected in the PCA plot, where networks segregate based on the database used (Figure \ref{fig:figure6}B).
  It is also evident that some networks, especially those inferred using the direct association network inference methods, are much closer in the PCA plot regardless of the reference database used.
  Additionally, the variation (on the first two principal axes) between the networks decreases when the low abundance \ac{otu}s are removed from the network (Figure~\ref{fig:figure_s1}).
  These results suggest that the most important criterion for accurate comparative analysis of co-occurrence networks is the taxonomy reference database followed by the level of filtering of the taxonomy tables.

@@

  The analysis of the rewiring of associations in the ASD samples with respect to the control provides a guide for identification of key genera that could be linked to dysbiosis.
  We observed 20 unique links in the network for control samples, 12 unique links in the network for ASD subjects and 7 edges in common between both networks.
  Although these unique connections do not imply real interactions, they can still serve as potential starting points for literature surveys and further experimental confirmations.
  For example, the changed connectivity of Prevotella and Porphyromonas in our networks show that these genera could be useful candidates for further downstream verification.
  Correspondingly, we have found previous studies that indicate that these genera are implicated in ASD~\cite{Kang2017,hoGutMicrobiotaChanges2020} and cognitive impairment~\cite{chiPorphyromonasGingivalisInducedCognitive2021}.

>>>>

  We developed \ac{micone}, a flexible and modular pipeline for the inference of co-occurrence networks from 16S data.
  \ac{micone} incorporates various popular, publicly available tools as well as custom Python modules for 16S data analysis and network inference (Methods).
  The different steps that are a part of the \ac{micone} co-occurrence network inference workflow (Figure~\ref{fig:figure1}) can be grouped into five major modules; (i) \ac{sp}; (ii) \ac{dc}; (iii) \ac{ta}; (iv) \ac{op}; and (v) \ac{ni}.
  Each process in the pipeline is implemented through multiple tools (see Methods and Figure~\ref{fig:figure1}).
  The effects of changing any intermediate step of the pipeline can be evaluated in terms of the final network outcome, as well as on any of the intermediate metrics and data outputs.
  The choice of tools and parameters is encoded in a configuration file (with parameters as shown in Tables S2-S6 at \href{https://github.com/segrelab/MiCoNE-pipeline-paper}{https://github.com/segrelab/MiCoNE-pipeline-paper}).
  Through a systematic analysis of tool combinations at each step of the pipeline, we estimated how much the final co-occurrence network depends on the possible choices at each step.

  Our analysis involved two types of data: The first type consisted of 16S sequencing data from samples of human stool microbiomes from a fecal microbiome transplant (FMT) study of autism~\cite{Kang2017}.
  The second type was a collection of datasets synthetically or artificially created for the specific goal of evaluating computational analysis tools.
  In particular, in order to benchmark each step in \ac{micone}, we used both mock data (labeled mock4, mock12, and mock16) from mockrobiota~\cite{Bokulich2016} and synthetic networks generated using the NorTA~\cite{Kurtz2015} and seqtime~\cite{Rottjers2018} approaches (See Methods).

@@

  \subsection*{DC: Denoising and clustering methods differ in their identification of sequences that are low in abundance}

  The \ac{dc} step is commonly carried out to generate representative sequences (in the form of the \acs{otu}/\acs{esv} tables) from the demultiplexed and trimmed 16S sequencing data.
  In order to compare the count tables generated by different tools, we processed the 16S sequencing reads (from the FMT study~\cite{Kang2017}) using 5 different methods: open-reference clustering, closed-reference clustering, de novo clustering, \ac{dada2}~\cite{Callahan2016} and Deblur~\cite{Amir2017}.
  The first three methods are from the vsearch plugin from \ac{qiime2}~\cite{bolyenReproducibleInteractiveScalable2019}.
  The closed and open reference methods in this analysis use the \acl{gg}~\cite{DeSantis2006} database for reference sequence alignment.

  A comparison of the different methods was carried out by calculating the mean UniFrac distances across all samples (Figure~\ref{fig:figure2}).
  The analysis was performed using both the weighted UniFrac~\cite{Lozupone2007} (Figure~\ref{fig:figure2}A) distance metric, which takes into account the counts of the representative sequences, and the unweighted UniFrac~\cite{Lozupone2005} (Figure~\ref{fig:figure2}B) distance metric, which gives equal weights to each sequence.

  The first main message emerging from this analysis is that the representative sequences generated by the different methods, with the exception of Deblur, are similar to each other when weighted by their abundance (Figure~\ref{fig:figure2}A).
  A second message is that the different methods differ mainly in the assignment of sequences of lower abundance.
  This can be inferred from the unweighted comparison (Figure~\ref{fig:figure2}B) which shows an increase in dissimilarity between each pair of methods (see additional details in Supplementary and Figure \ref{fig:figure_s2}).

  These comparisons only elucidate the similarity between a pair of methods.
  To determine which tool most accurately recapitulates the reference sequences in the samples, we applied the same pipeline step to process the mock datasets (mock4, mock12, and mock16) and compared the predicted representative sequences with the true sequences and their distribution.
  The results (Figure~\ref{fig:figure2}C and \ref{fig:figure2}D) show that the predicted sequence distributions are overall different from the expected ones.
  The variation across datasets indicates that the datasets themselves play a big role in method performance.
  There is no method that outperforms the rest in all datasets (see Supplementary for an extended discussion).
  Based on their slightly better performance on the mock datasets, their de novo error correcting nature and previous independent evaluation~\cite{Nearing2018}, \ac{dada2} and Deblur appear to be the most reliable.
  This is because the open-reference and de novo clustering methods return a much larger number of \ac{otu}s compared to the other pipelines and would affect the accuracy of the network inference step if stringent filtering is not performed.
  Overall, since \ac{dada2} as compared to Deblur, displays better performance on all the mock datasets on the weighted UniFrac metric, we set this tool as the default for the DC step of the pipeline.
  However, if comparison across studies that have sequenced different 16S regions is required, closed-reference and open-reference might be a better option.

  After the denoising, the sequences are subject to Chimera Checking (CC).
  The \ac{micone} pipeline supports two different chimera checking methods, ``uchime-denovo"~\cite{bolyenReproducibleInteractiveScalable2019}, and ``remove bimera"~\cite{Callahan2016}.
  We did not notice any notable difference between the two methods (Figure~\ref{fig:figure_s3}), implying that they identify and remove mostly the same set of sequences as chimeras.
  Since the remove bimera method was originally developed in conjunction with dada2 we use this method as the default.
  The DC step thus results in a reduced set of unique sequences, which will be referred to as representative sequences in the subsequent steps.

@@

  After assigning all the representative sequences to taxonomies we performed a pairwise comparison of the similarity between the top 100 assignments (by abundance) from different databases at every taxonomic level (Figure~\ref{fig:figure3}B).
  The comparisons of the assignments beyond the Family level (Family, Genus, and Species) show less than $45\%$ similarity between any pair of databases.
  This implies that the taxonomy assignments from each reference database are fairly unique.
  The comparison of all assigned genera (Figure~\ref{fig:figure_s4}), instead of just the top 100, contains a higher percentage of mismatches.
  This suggests that many of the most abundant sequences are consistently matched to the same taxonomies in the different reference databases, at least for the dataset tested in the current analysis.

@@

  Since no database performs better than others against mock datasets, the choice of which database to use could be driven by other reasons (see Supplementary discussion).
  One reason to choose a particular database could be the frequency of updates and the potential for future growth.
  Both \ac{gg}, due to its frequent use in the literature~\cite{Balvociute2017}, and \ac{ncbi}, due to its regular revision and maintenance, could be good choices for taxonomy assignment.
  In our default pipeline, we choose \ac{gg} as the default method.

@@

  Similar to what was done for the previous steps of the pipeline, and in analogy with previous estimations of network inference accuracy~\cite{Kurtz2015,Weiss2016}, we evaluated the network inference algorithms and the final consensus network using synthetic interaction data.
  For this purpose, we generated synthetic interaction data using the ``NorTA''~\cite{Kurtz2015} and ``seqtime''~\cite{faustSignaturesEcologicalProcesses2018} methods (see Methods).
  For each method, an \ac{otu} counts table is generated based on the selected interaction network topology and properties.
  This counts table is used as the input to the \ac{micone} pipeline to generate predicted associations.
  The interaction network used to generate the counts table is used as the source of true interactions to calculate the precision (Figure \ref{fig:figure5}) and sensitivity (Figure \ref{fig:figure_s5} and Figure \ref{fig:figure_s6}) for each network inference algorithm.
  As shown in Figure \ref{fig:figure5} the consensus algorithm, especially the scaled-sum method, captures true associations with high precision (through the removal of edges that are either not present in most of the inference methods or whose association strength is low across methods).
  Overall, the scaled-sum method for $p=1.000$ performs the best (precision = $1.000$ for both NorTA and seqtime).
  The scaled-sum method for $p=0.333$ (default option in the pipeline) shows a high precision ($0.956$ with NorTA; $0.688$ with seqtime), without displaying significant reduction in sensitivity (Figure~\ref{fig:figure_s5} and Figure~\ref{fig:figure_s6}).
  However, if higher precision is required $p>0.5$ can be considered.
  Therefore, the consensus networks provide the means to obtain a short list of associations that would have a high probability of being present in the real association network.

@@

  In order to analyze the effect of different processing methods on the inferred co-occurrence networks (before consensus estimation), we generated networks using all possible combinations of methods and quantified the variability due to each choice (Figure \ref{fig:figure6}A).
 This was achieved by building a linear model of the edges of the network as a function of the various steps in the pipeline workflow (see Methods).
  Figure \ref{fig:figure6}A, shows the percentage of total variation among the co-occurrence networks due to the different steps of the pipeline.
  The \ac{ta} step, or more specifically the choice of 16S reference database, contributes the most ($\sim65.4\%$) to the variation in the networks, followed by the \ac{op} step ($\sim26.8\%$).
  This result highlights the importance of the taxonomy assignment step in the 16S data analysis workflow, implying that a change in the reference database will result in drastically different inferred networks.
  This is likely due to the differential assignment of representative sequences to taxonomic entities (Figure~\ref{fig:figure3} and Figure~\ref{fig:figure_s4}), which drastically alter the nodes and hence the underlying network topology.

  The effects of the different steps of the pipeline on the inferred networks can be best visualized through dimensionality reduction.
  The PCA in Figure \ref{fig:figure6}B shows all the above networks, colored by the databases used in TA (left) and by the NI approach (right).
  The major effect of the TA step choice, shown before in Figure \ref{fig:figure6}A, is also reflected in the PCA plot, where networks segregate based on the database used (Figure~\ref{fig:figure6}B and Figure~\ref{fig:figure_s1}).
  It is also evident that some networks, especially those inferred using the direct association network inference methods, are much closer in the PCA plot regardless of the reference database used.
  Additionally, the variation (on the first two principal axes) between the networks decreases when the low abundance \ac{otu}s are removed from the network.
  These results suggest that the most important criterion for accurate comparative analysis of co-occurrence networks is the taxonomy reference database followed by the level of filtering of the taxonomy tables.

@@

  The analysis of the rewiring of associations in the ASD samples with respect to the control provides a guide for the identification of key genera that could be linked to dysbiosis.
  We observed 20 unique links in the network for control samples, 12 unique links in the network for ASD subjects, and 7 edges in common between both networks.
  Although these unique connections do not imply real interactions, they can still serve as potential starting points for literature surveys and further experimental confirmations.
  For example, the changed connectivity of Prevotella and Porphyromonas in our networks show that these genera could be useful candidates for further downstream verification.
  Correspondingly, we have found previous studies that indicate that these genera are implicated in ASD~\cite{Kang2017,hoGutMicrobiotaChanges2020} and cognitive impairment~\cite{chiPorphyromonasGingivalisInducedCognitive2021}.
  Additional visualization and comparison of networks can be performed using the \acl{mind}~\cite{huResourceComparisonIntegration2022}.
